signalHsmm - a novel semi-Markov model of eukaryotic signal peptides

### 2015-06-16 paper.tex
adding model schemes

Hidden Markov Model(HMM) is a widely used machine learning technique which, in contrast to black-box methods like neural network used by SignalP, 
is easy to interpret and allows researcher to analyze and understand structure of modeled phenomenon. 
In case of signal-peptide, we might wish to verify whether regions really describe it. Also we might capture
structure for some special subgroups and with HMM we can verify biological hypothesis.

Suppose you have a sequence of observations e.g. amino acids. What is of interest is
understanding underlying cause of their occurrence. 
HMM aims to answer that question by assuming a specific yet flexible structure of the problem.
HMM consists of two stochastic processes. First is discrete Markov chain $X_{t=1}^T$ on the set of so called hidden states $\{S_1, \dots S_n\}$.
These are "the cause" for the observations. At every step, hidden state might change according to transition matrix
$A= (a)_{i,j=1}^n$, where $a_{i,j} = \mathcal{P}(X_{t+1} = S_j | X_t = S_i)$. In our application hidden states are signal peptide regions.
Second process $E_{t=1}^T$ is observation process, defined on the set of possible observations $\{O_1, \dots, O_m\}$. 
Observations are assumed to occur independently. Their distribution depends only on hidden state which emits it. 
Probabilities are given by a matrix $B$, $b_{i,k} = \mathcal{P}(E_t = O_k | X_t = S_i)$.
In our case, observations are (degenerated) amino acids.
Main goal in signalHsmm is to find, for a given peptide, most probable regions boundaries. This is achieved with
Viterbi algorithm.
For a good reference on HMM see~\citep{1989rabinera}.

In regular HMM, hidden state duration -- number of observations emitted by the hidden state, has geometric distribution.
\citep{Durbin98biologicalsequence} shows how to extend it for some different distributions without significant increase in computational 
complexity. Similar ideas were used for signal peptide recognition, for example in~\citep{2004klla}. 
However it still not flexible enough. Empirical regions length distributions (see Figure~\ref{fig:reglen}),
is difficult to capture in that way.

@@

The gain is not only a better fit to the data but also, as number of hidden states is smaller, easier interpretation 
for researcher without mathematical background.

HSMM can be further extend to use some specific knowledge of motives - n-grams.

>>>>

Let us first define Hidden Markov Model (HMM). Suppose you have a sequence of observations e.g. amino acids. What is of interest is
understanding underlying cause of their occurrence. 
HMM aims to answer that by assuming a specific yet flexible structure of the problem.
HMM consists of two stochastic processes. First is discrete Markov chain $X_{t=1}^T$ on the set of so called hidden states $\{S_1, \dots S_n\}$.
These are "the cause" for the observations. At every step, hidden state might change according to transition matrix
$A= (a)_{i,j=1}^n$, where $a_{i,j} = \mathcal{P}(X_{t+1} = S_j | X_t = S_i)$. In our application hidden states are signal peptide regions.
Second process $E_{t=1}^T$ is observation process, defined on the set of possible observations $\{O_1, \dots, O_m\}$. 
Observations are assumed to occur independently. Their distribution depends only on hidden state which emits it. 
Probabilities are given by a matrix $B$, $b_{i,k} = \mathcal{P}(E_t = O_k | X_t = S_i)$.
In our case, observations are (degenerated) amino acids.
Main goal in signalHsmm is to find, for a given peptide, most probable regions boundaries. This is achieved with
Viterbi algorithm.
For a good reference on HMM see~\citep{1989rabinera}.

In regular HMM, hidden state duration -- number of observations emitted by the hidden state, has geometric distribution.
\citep{Durbin98biologicalsequence} shows how to extend it for some different distributions without significant increase in computational 
complexity. Similar ideas were used for signal peptide recognition, for example in~\citep{2004klla}. 
It is However still not flexible enough. Empirical regions length distributions (see figure~\ref{fig:reglen})
are difficult to capture in that way.

@@

The gain is not only a better fit to the data but also, as 
model structure (see figure \ref{fig:model_hsmm}) is straightforward, easier interpretation 
for researcher without mathematical background.

%model structure
\begin{figure}
 \begin{minipage}{.5\linewidth} 
 \caption{Scheme of signalHsmm}
     \resizebox{6.5cm}{!}{%
    \begin{tikzpicture}[->,>=stealth',shorten >=2pt,auto,node distance=4.5cm, thick]
    \label{fig:model_hsmm}
      \tikzstyle{line} = [draw=black, color=blue!30!black!50, line width=4.5mm, -latex']
      \tikzstyle{main node} = [circle,fill=blue!20,draw, minimum size = 2.2cm, font=\itshape,
         align=center,  top color=white, bottom color=blue!50!black!70 ] %font=\sffamily\small\bfseries,
      %nodes
      \node[main node]            (start') 	[]						{Start};	     
      \node[main node, bottom color=purple!70!black!70] 	(nregion') 	[right of=start',xshift=-5mm, yshift=15mm] 	{n-region};
      \node[main node, bottom color=pink!70!black!70] 	(hregion') 	[right of=nregion',xshift=-5mm,yshift=15mm] 	{h-region};
      \node[main node, bottom color=gray!70!black!70] 	(cregion') 	[right of=hregion',xshift=-5mm,yshift=-15mm] 	{c-region};
      \node[main node, bottom color=green!70!black!70] 	(mature') 	[right of=cregion',xshift=-5mm, yshift=-15mm] 	{Mature protein};

      %lines
      \path [line] (start')   edge node [left, color=black] {} (nregion');
      \path [line] (nregion') edge node [below, color=black] { } (hregion');
      \path [line] (hregion') edge node [below, color=black] { } (cregion');
      \path [line] (cregion') edge node [left, color=black] { } (mature');
      \draw [line] (start') to[out=340,in=200] (mature');
    \end{tikzpicture} }
\end{minipage}
 \begin{minipage}{.5\linewidth}
   \caption{Scheme of signalHsmm with cleavage site motives}
     \resizebox{6.5cm}{!}{%
    \begin{tikzpicture}[->,>=stealth',shorten >=2pt,auto,node distance=4.5cm, thick]
    \label{fig:model_hsmm_kMer}
      \tikzstyle{line} = [draw=black, color=blue!30!black!50, line width=4.5mm, -latex']
      \tikzstyle{main node} = [circle,fill=blue!20,draw, minimum size = 2.2cm, font=\itshape,
         align=center,  top color=white, bottom color=blue!50!black!70 ] %font=\sffamily\small\bfseries,
      %nodes
      \node[main node]            (start')   []  					{Start};	     
      \node[main node, bottom color=purple!70!black!70] 	(nregion') 	[right of=start',xshift=-25mm, yshift=35mm] 	{n-region};
      \node[main node, bottom color=pink!70!black!70] 	(hregion') 	[right of=nregion',xshift=-25mm,yshift=35mm] 	{h-region};
      \node[main node, bottom color=gray!70!black!70] 	(cregion') 	[right of=hregion',xshift=-10mm,yshift=-25mm] 	{c-region};
      \node[main node, bottom color=green!70!black!70] 	(mature') 	[below of=hregion',xshift=15mm, yshift=-90mm] 	{Mature protein};

      \node[main node, bottom color=yellow!70!black!70]   (kmer1) 	[below of=cregion',xshift=35mm,yshift=30mm] 	{Aminoacid 1};
      \node[main node, bottom color=yellow!70!black!70]   (kmer2) 	[below of=kmer1,xshift=-5mm,yshift=5mm] 	{Aminoacid 2};
      \node[main node, bottom color=yellow!70!black!70]   (kmer3) 	[below of=kmer2,xshift=-5mm,yshift=5mm] 	{Aminoacid 3};
      %lines
      \path [line] (start')   edge node [left, color=black] {} (nregion');
      \path [line] (nregion') edge node [below, color=black] { } (hregion');
      \path [line] (hregion') edge node [below, color=black] { } (cregion');
      \path [line] (cregion') to[out=230,in=90] (mature');
      \draw [line] (start') to[out=300,in=180] (mature');

      \draw [line] (cregion') edge node [below, color=black] { } (kmer1);
      \draw [line] (kmer1) edge node [left, color=black] {} (kmer2);
      \draw [line] (kmer2) edge node [left, color=black] {} (kmer3);
      \draw [line] (kmer3) edge node [left, color=black] {} (mature');

      %\draw [decorate,line width=3pt,decoration={brace,amplitude=10pt}] ([xshift=5mm, yshift=5mm]kmer1) -- ([yshift=-25mm]kmer3) node [black,midway,xshift=0.6cm] {\footnotesize $P_1$};
      \draw[decoration={brace,amplitude=10pt, raise=5pt},decorate, line width=2pt]  
      ([yshift=1mm]kmer1.east) -- ([yshift=-5mm]kmer3.east) node [black,midway,xshift=0.6cm] {\large \color{yellow!50!black!90} k-mer};
    \end{tikzpicture} }
    \end{minipage}
\end{figure}

HSMM can be further extend to use some specific motives that might occur in proximity of cleavage site.
Such an extension requires defining additional hidden states. 
It proves to be extremely useful in recognizing cleavage site.
Scheme can be seen in figure \ref{fig:model_hsmm_kMer}.
